CrystalPlan: an Experiment Planning Tool for Crystallography

### 2011-02-23 docs/CrystalPlan_paper/paper.tex
Crystal plan paper, revision 2 submitted to journal.

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Laboratory (ORNL), currently finishing commissioning, is a neutron
time-of-flight Laue diffractometer with high wavelength (time) resolution. It
is designed to acquire single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering three
quarters of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a quarter of $2\pi$ steradian coverage. 
To have sufficient data for a reliable structure solution and
refinement, an experimenter will typically want to measure $ > 85\%$ of the
unique reflections within a certain range of d-spacings. However, TOPAZ's wide
wavelength band combined with its complex geometry makes it difficult to come up with
an effective set of sample orientations that will cover all the required reflections
without unnecessary redundancy.

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ. With some relatively minor modifications, the program
could also be adapted for use at reactor sources. Monochromatic reactor sources
could be handled by adding support sweeps consisting of many motor positions;
Laue diffractometers would requiring detecting and removing reflections with chromatic overlap.

@@

\begin{figure}
\caption{Sample screenshot showing the 3D view of the predicted coverage of
reciprocal space volume for 3 sample orientations and 14 detectors on TOPAZ,
from 0.9 \ang\ to 4.0\ang: a) without crystal symmetry, 17\% coverage; b) with crystal symmetry (mmm 8-fold
symmetry), coverage increases to 65\%. }
\ifdograph
  \includegraphics{volume_nosym}
\fi
a)
\ifdograph
  \includegraphics{volume_sym}
\fi
b)
\label{fig:volume_view} 
\end{figure}

>>>>

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Laboratory (ORNL), currently finishing commissioning, is a neutron
time-of-flight Laue diffractometer with high wavelength (time) resolution. It
is designed to acquire single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering
approximately half of $4\pi$ steradians. However as of this writing, only 14 of
the 48 detectors are installed, giving only a sixth of $4\pi$ steradian
coverage. To have sufficient data for a reliable structure solution and  
refinement, an experimenter will typically want to measure $ > 85\%$ of the
unique reflections within a certain range of d-spacings. However, TOPAZ's wide 
wavelength band combined with its complex geometry makes it difficult to come up
with  an effective set of sample orientations that will cover all the required
reflections without unnecessary redundancy.

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ. With some relatively minor modifications, the program
could also be adapted for use at reactor sources. Monochromatic diffractometers
could be handled by adding support sweeps consisting of many motor positions;
Laue diffractometers on a constant intensity beam would requiring
detecting and removing reflections with chromatic overlap.

@@

\begin{figure}
\caption{Sample screenshot showing the 3D view of the predicted coverage of
reciprocal space volume for 3 sample orientations and 14 detectors on TOPAZ,
from 0.9 \ang\ to 4.0\ang: a) without crystal symmetry, 17\% coverage; b) with crystal symmetry (mmm 8-fold
symmetry), coverage increases to 65\%. The area plot at the bottom shows the
measured percentage of reciprocal volume as a function of $q$; the
color indicates the redundancy (cyan: 1, green: 2, yellow: 3, orange: 4). }
\ifdograph
  \includegraphics{volume_nosym} 
\fi
a)
\ifdograph
  \includegraphics{volume_sym}
\fi
b)
\label{fig:volume_view} 
\end{figure}
